---
title: "Bellabeat Case Study with R"
author: "Henrique Almeida"
date: "13/01/2022"
output: html_document
---

This project has been done as a capstone project for the Google Data Analytics course, completed by me at Coursera platform. The goal is to analyze a public data from fitness tracker smart devices in order to support the marketing strategy of Bellabeat, a high-tech manufacturer of health-focused products for women.

## Scenario
For this project, I'm playing the role of a junior data analyst working on the marketing analyst team at Bellabeat, a high-tech manufacturer of health-focused products for women. Bellabeat is a successful small company, but they have the potential to become a larger player in theglobal smart device market. Urška Sršen, cofounder and Chief Creative Officer of Bellabeat, believes that analyzing smart device fitness data could help unlock new growth opportunities for the company. You have been asked to focus on one of Bellabeat’s products and analyze smart device data to gain insight into how consumers are using their smart devices. The insights you discover will then help guide marketing strategy for the company.


## Phase 1: Ask

Sršen asks you to analyze smart device usage data in order to gain insight into how consumers use non-Bellabeat smart devices. She then wants you to select one Bellabeat product to apply these insights to in your presentation. These questions
will guide the analysis:

1. What are some trends in smart device usage?
2. How could these trends apply to Bellabeat customers?
3. How could these trends help influence Bellabeat marketing strategy?

To better structure the analysis, I'll be answering to some questions.

What is the problem you are trying to solve?

- Gain insights on non-Bellabeat smart device’s usage through data analysis in order to drive recommendations to support Bellabeat’s marketing strategy.

How can your insights drive business decisions?

- If we can understand how the target audience uses health trackers and smart devices from other companies, we can draft marketing strategies that focus on the most using patterns, common situations, main pain points, and so on.

## Phase 2: Prepare


The data contains info about:

- Activity
- Calories
- Intensities
- Steps
- Sleep pattern
- weight 


Installing the `tidyverse` package to analyze the data:

 ```{r}
 install.packages("tidyverse")
 ```

Loading the `tidyverse` package:
```{r}
library(tidyverse)
```

Loading the data: 
```{r}
daily_activity <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
daily_calories <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyCalories_merged.csv")
daily_intensities <- read_csv("Fitabase Data 4.12.16-5.12.16/dailyIntensities_merged.csv")
daily_steps <- read_csv("Fitabase Data 4.12.16-5.12.16/dailySteps_merged.csv")
sleep_day <- read_csv("Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
weight_info <- read_csv("Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")

```



